---
title: "datacode"
output:
  html_document: default
  word_document: default
  pdf_document: default
date: "2026-06-29"
---
##Step 0: packages and data
```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(emmeans)


data <- read_csv("ubelong_longitudinalData_FINAL201225_forReShare.csv",
                 show_col_types = FALSE)
```

## step1 check nesting

```{r cars}

data %>%
  count(timepoint)


data %>%
  count(participant_id) %>%
  count(n)


data %>%
  summarise(
    n_rows = n(),
    n_participants = n_distinct(participant_id),
    missing_ucla = sum(is.na(UCLA_freq)),
    missing_belonging = sum(is.na(BES_full_score)),
    missing_timepoint = sum(is.na(timepoint))
  )
```
```{r}
data %>%
  count(participant_id) %>%
  count(n)
```
```{r}
data %>%
  group_by(timepoint) %>%
  summarise(
    n = n(),
    missing_ucla = sum(is.na(UCLA_freq)),
    missing_ucla_pct = mean(is.na(UCLA_freq)) * 100,
    missing_belonging = sum(is.na(BES_full_score)),
    missing_belonging_pct = mean(is.na(BES_full_score)) * 100
  )
```
##Step 1.5：检查 T2 缺失是不是系统性的
```{r}
t2_status <- data %>%
  select(participant_id, timepoint, UCLA_freq, BES_full_score) %>%
  mutate(
    ucla_observed = !is.na(UCLA_freq),
    bes_observed = !is.na(BES_full_score)
  ) %>%
  group_by(participant_id) %>%
  summarise(
    t2_ucla_observed = ucla_observed[timepoint == "T2"],
    t2_bes_observed = bes_observed[timepoint == "T2"],
    .groups = "drop"
  )

t1_attrition <- data %>%
  filter(timepoint == "T1") %>%
  left_join(t2_status, by = "participant_id")
```
time2：
```{r}


t1_attrition %>%
  group_by(t2_ucla_observed) %>%
  summarise(
    n = n(),
    mean_t1_ucla = mean(UCLA_freq, na.rm = TRUE),
    sd_t1_ucla = sd(UCLA_freq, na.rm = TRUE),
    mean_t1_belonging = mean(BES_full_score, na.rm = TRUE),
    sd_t1_belonging = sd(BES_full_score, na.rm = TRUE)
  )
```
```{r}
t.test(UCLA_freq ~ t2_ucla_observed, data = t1_attrition)
t.test(BES_full_score ~ t2_ucla_observed, data = t1_attrition)
```

##Step 2: define analysis sample
```{r}
data_analysis <- data %>%
  mutate(
    participant_id = factor(participant_id),
    timepoint_f = factor(timepoint, levels = c("T1", "T2")),
    time_c = if_else(timepoint == "T2", 1, 0),
    surveyYear_f = factor(surveyYear),
    sex_f = factor(sex),
    yearOfStudy_f = factor(yearOfStudy)
  ) %>%
  select(
    participant_id,
    timepoint,
    timepoint_f,
    time_c,
    UCLA_freq,
    BES_full_score,
    surveyYear_f,
    sex_f,
    yearOfStudy_f
  ) %>%
  drop_na()

```
检查模型样本
```{r}
data_analysis %>%
  summarise(
    n_obs = n(),
    n_participants = n_distinct(participant_id)
  )

data_analysis %>%
  count(timepoint_f)

data_analysis %>%
  count(participant_id) %>%
  count(n)
```
##step3：centering
```{r}
data_analysis <- data_analysis %>%
  group_by(participant_id) %>%
  mutate(
    belonging_person_mean = mean(BES_full_score),
    belonging_within = BES_full_score - belonging_person_mean
  ) %>%
  ungroup() %>%
  mutate(
    belonging_between = as.numeric(scale(
      belonging_person_mean,
      center = TRUE,
      scale = FALSE
    ))
  )
```
##model0：nonmodel
```{r}
m0 <- lmer(
  UCLA_freq ~ 1 + (1 | participant_id),
  data = data_analysis,
  REML = FALSE
)

summary(m0)
performance::icc(m0)
```

##model1:time effect

```{r}
m1 <- lmer(
  UCLA_freq ~ time_c + (1 | participant_id),
  data = data_analysis,
  REML = FALSE
)

summary(m1)
```

##Model 2: 加 within-person belonging

```{r}
m2 <- lmer(
  UCLA_freq ~ time_c + belonging_within +
    (1 | participant_id),
  data = data_analysis,
  REML = FALSE
)

summary(m2)
```
##model3：between person belonging
```{r}
m3 <- lmer(
  UCLA_freq ~ time_c + belonging_within + belonging_between +
    (1 | participant_id),
  data = data_analysis,
  REML = FALSE
)

summary(m3)
```

##model4 covirates
```{r}
m4 <- lmer(
  UCLA_freq ~ time_c + belonging_within + belonging_between +
    surveyYear_f + sex_f + yearOfStudy_f +
    (1 | participant_id),
  data = data_analysis,
  REML = FALSE
)

summary(m4)
```

##Model comparison
```{r}
anova(m0, m1, m2, m3, m4)
AIC(m0, m1, m2, m3, m4)
BIC(m0, m1, m2, m3, m4)
```
##random slope check

m4_rs <- lmer(
  UCLA_freq ~ time_c + belonging_within + belonging_between +
    surveyYear_f + sex_f + yearOfStudy_f +
    (1 + time_c | participant_id),
  data = data_analysis,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

summary(m4_rs)
isSingular(m4_rs)

Random slopes for time were considered but could not be estimated because the number of random effects exceeded the number of observations. Therefore, the simpler random-intercept model was retained.

##model4 as retained model
```{r}
m4 <- lmer(
  UCLA_freq ~ time_c + belonging_within + belonging_between +
    surveyYear_f + sex_f + yearOfStudy_f +
    (1 | participant_id),
  data = data_analysis,
  REML = FALSE
)
```
##final REML model
```{r}
m_final <- lmer(
  UCLA_freq ~ time_c + belonging_within + belonging_between +
    surveyYear_f + sex_f + yearOfStudy_f +
    (1 | participant_id),
  data = data_analysis,
  REML = TRUE
)

summary(m_final)
performance::icc(m_final)
performance::r2_nakagawa(m_final)
```
用 time、within-person belonging、between-person belonging 和 covariates 预测 UCLA loneliness，并允许每个 participant 有自己的 random intercept。

也就是：

fixed effects：平均效应；
random intercept：每个学生有自己的平均 loneliness 水平；
没有 random slope：因为数据不支持估计 individual change slope。

###A. Assumption checks
```{r}
# Residuals vs fitted: check homoscedasticity / nonlinearity
plot(fitted(m_final), resid(m_final))
abline(h = 0, lty = 2)

# Normality of level-1 residuals
qqnorm(resid(m_final))
qqline(resid(m_final))

# Normality of random intercepts
ranef_values <- ranef(m_final)$participant_id[, "(Intercept)"]
qqnorm(ranef_values)
qqline(ranef_values)
```
###B. Complete two-wave sensitivity analysis（for t2 missing）
```{r}
complete_ids <- data_analysis %>%
  count(participant_id) %>%
  filter(n == 2) %>%
  pull(participant_id)

data_complete2 <- data_analysis %>%
  filter(participant_id %in% complete_ids)

m_final_complete2 <- lmer(
  UCLA_freq ~ time_c + belonging_within + belonging_between +
    surveyYear_f + sex_f + yearOfStudy_f +
    (1 | participant_id),
  data = data_complete2,
  REML = TRUE
)

summary(m_final_complete2)
performance::r2_nakagawa(m_final_complete2)
```

